openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2022-04-08 paper/main.tex
Finish up implementation section

$t$-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method that to aims to find a low-dimensional embedding
where neighborhoods are preserved. More formally, given a multi-dimensional data set
$\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \}
\in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE
aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1,
\mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$,
such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the
high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and
$\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization
tool, $d$ is typically set to two. The similarity between two data points in the
high-dimensional space is defined using the Gaussian kernel
\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0 \notag
\label{eq:gaussian_kernel}
\end{equation}
where $\mathcal{D}$ is some distance measure. The values $p_{j \mid i}$
are then symmetrized to
\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

@@

We introduce \opentsne, a comprehensive Python library that implements the
t-SNE algorithm and its numerous recentl extensions. \opentsne\ is
designed to accomodate all types of users and includes both a simple API that
allows users to quickly obtain informative t-SNE visualizations, as well as a
more flexible API that allows seasoned practitioners to modify all the
different aspects of the t-SNE algorithm.

The t-SNE algorithm is comprised of three main steps, mirrored in the structure
and flow of the \opentsne\ library. These are: the initial positions
of points in the embedding, an affinity model describing the similarities
between data points in the embedding, and the optimization phase which
optimizes the embedding layout. Using the advanced API, each of these key
elements can be modified or tuned according to produce the best possible
visualization.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{opentsne-diagram}
  \caption{\label{fig:diagram}
  Bla
  }
\end{figure*}

\opentsne\ is compatible with the Python data science ecosystem (e.g. \pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar and intuitive API to end-users. Its modular design encourages extensibility and experimentation with various parameter settings and changes to the analysis pipeline. \opentsne\ aims to make the latest theoretical advances in t-SNE accessible to the wider data science community. These features are easily accessible through an intuitive API, which closely follows the style \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit in the machine learning community.
The following snippet exemplifies this interface and demonstrates how \opentsne\ can be used to quickly generate several visualizations, each highlighting different aspects of the underlying data structure.

\subsection{Example}

@@

\opentsne\ is designed to be as user-friendly as possible and can be installed through either \pkg{PyPI} using \pkg{pip}
\begin{CodeChunk}
\begin{CodeInput}
pip install opentsne
\end{CodeInput}
\end{CodeChunk}
or using \pkg{conda} from the \pkg{conda-forge} channel
\begin{CodeChunk}
\begin{CodeInput}
conda install -c conda-forge opentsne
\end{CodeInput}
\end{CodeChunk}
\opentsne\ provides precompiled binaries for all major platforms (Windows, MacOS, Linux) for all currently supported Python versions (3.6---3.10). This is especially convenient for Windows users, as this obviates the need for a C/C++ compiler, which do not come bundled with the operating system.

@@

We demonstrate how to use \opentsne\ on the Iris data set from \citet{anderson1936species}.
This small data set describes 150 iris flowers containing four variables (sepal
length, sepal width, petal length, petal width) and their species (Iris Setosa,
Iris Versicolor, Iris Virginica), which is the target variable.
This data set is included in \pkg{scikit-learn}, making it easy to load:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
\pkg{scikit-learn} loads the data set into a \pkg{numpy} array, which can directly
be used by \opentsne:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
>>>
>>> embedding = openTSNE.TSNE().fit(iris.data)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
\opentsne\ returns an \code{TSNEEmbedding} object, which is a subclass of
\code{np.ndarray}. This allows users to easily manipulate the resulting embedding.
The additional functionality provided by the \code{TSNEEmbedding} object allows
the embedding to be further optimized using different parameter values, or be
used as a reference when obtaining new data points.
For users who do not need this additional functionality, this object can be
converted to a standard \code{np.ndarray} with:
\begin{CodeChunk}
\begin{CodeInput}
>>> import numpy as np
>>>
>>> X_tsne = embedding.view(np.ndarray)
>>> X_tsne[:3]
array([[-0.55581917, 17.56162344],
       [-1.76742536, 20.07763453],
       [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
Fig.~\ref{fig:iris} shows a visualization of the resulting t-SNE visualization.
As expected, t-SNE is correctly able to group together samples from the same
species. Additionally, the t-SNE embedding reveals that Iris Setosa is markedly
different from the other two species, while flowers from Iris Versicolor and
Iris Virginica may appear more similar to one another.
\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{iris}
  \caption{\label{fig:iris}
  Visualization of the obtained t-SNE embedding on the Iris data set. The colors
  correspond to different species of the Iris flower.
}
\end{figure*}

\section{Discussion and Examples} \label{sec:summary}

@@

We illustrate this point by generating four different embeddings of the data on single-cell gene expression in mouse brain~\citep{tasic2018shared}. Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
We generate the four different embeddings using the following code snippet (which is shortened for clarity).
\begin{CodeChunk}
\begin{CodeInput}
embedding_a = TSNE(perplexity=30).fit(x)
embedding_b = TSNE(perplexity=500).fit(x)
embedding_c = TSNE(perplexity=[30, 500]).fit(x)
embedding_d = TSNE(perplexity=30, dof=0.6).fit(x)
\end{CodeInput}
\end{CodeChunk}
While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear equidistant from their neighbors, and the overall relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent. Here, it is evident from the increased white space between groups that there is one large class of excitatory neurons and two related classes of inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods leads to the absorption of smaller clusters into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve both the global organization of clusters and prevent smaller cluster absorption. We constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. The figure demonstrates that some clusters are composed of numerous, smaller subgroups representing different cell subpopulations that are not visible under standard parameter settings.

@@

When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development. The embedding indicates numerous clusters with transitions between time points, as marked by the color-coding, that are difficult to interpret. Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently, more informative visualizations~\citep{kobak2019art}. For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure. Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters. The visualization also reveals several tiny clusters, possibly corresponding to rare cell types.
\begin{CodeChunk}
\begin{CodeInput}
# To improve the global structure of the final layout, we initialize the
# embedding by creating a t-SNE plot on a sample of points

@@

# Create a t-SNE embedding on the sample, using a high perplexity value to
# emphasize the high level structure
init_sample = TSNE(perplexity=500).fit(x_sample)

@@

# Lastly, we merge `init_sample` and `init_rest` and restore the original order
init_full = np.vstack((init_sample, init_rest))[reverse]
\end{CodeInput}
\end{CodeChunk}
We generate the three different embeddings using the following code snippet (which is shortened for clarity).
\begin{CodeChunk}
\begin{CodeInput}
# We separately precompute the affinities as this can take quite a long time.
# This way, we can reuse the precomputed affinities in various different
# embeddings, without having to repeat the same, costly computation every time.
affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)

@@

Exaggeration can highlight transitions between cell states in developmental
studies. Standard t-SNE often produces embeddings with clearly defined, discrete
clusters. We can adjust the level of granularity and resolution of the clusters
with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
often undesired in developmental studies where cells' state is assumed to follow
a continuous transition path. To this end, other embedding techniques such as
UMAP and ForceAtlas2 are used to better capture the continuity between cell
states. Recently, \citet{bohm2020unifying} showed that
embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$
construct embeddings which are markedly similar to UMAP and ForceAtlas2,
respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory
between different time points is difficult to observe due to many sprawled out
clusters. On the other hand, it is easier to trace the development when we
increase the exaggeration factor from $1$ to $2$ to $4$ in
Figs.~\ref{fig:cao}.b-c.

@@

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
previously labeled single-cell data set and embed cells from a separate
experiment into the reference landscape. The reference data from
\citet{macosko2015highly} contains gene expression
profiles from mouse retinal cells. By embedding the samples from a similar
experiment on bipolar retinal cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar cell
clusters onto the reference embedding. The following snippet shows the code necessary to generate figure~\ref{fig:macosko}.
\begin{CodeChunk}
\begin{CodeInput}
# We specify the old default parameter values which are often used to show
# why t-SNE misbehaves or fails to preserve global structure
tsne_old = TSNE(initialization="random", learning_rate=200, n_iter=750).fit(x)

# By default, openTSNE uses PCA-based initialization and automatically
# determines the best learning rate. This also allows us to lower the default
# number of iterations to 500, lowering the overall runtime
tsne_standard = TSNE(perplexity=[50, 500]).fit(x)

@@

Versatility, the ability to use and combine different optimization approaches to
construct different embedding spaces, is another of \opentsne's core design
principles. \citet{kobak2019art} recently provided several recommendations and
tricks to obtain better and more meaningful t-SNE visualizations. These include
multi-scale similarity kernels, perplexity annealing, and increasing
exaggeration when working with massive data sets. \opentsne\ provides a flexible
API to incorporate these improvements in just a few lines of code. Furthermore,
\opentsne\ supports custom affinity models, enabling users to construct t-SNE
embeddings on non-tabular relational data: the only requirement imposed by the
affinity-model is some notion of similarity between data points. Finally,
\opentsne's comprehensive callback system can be utilized to monitor and adapt
different stages of the optimization phase and has been used to construct
visually appealing animations of the t-SNE optimization process.

@@

% General problem statement
One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data points~\citep{becht2019dimensionality}.
The culprit for slow response time stems from particular optimization procedures
and their specific implementation in popular
open-source libraries. Until quite recently, most popular implementations of
t-SNE were based on either a direct implementation of the t-SNE algorithm with
asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation
scheme developed by \citet{van2014accelerating} with asymptotic time complexity
$\mathcal{O}(N \log N)$. Recently, however, \citet{linderman2019fast} developed
a novel approximation scheme -- FIt-SNE -- which reduces the asymptotic time
complexity to $\mathcal{O}(N)$, overcoming the difficulties of scaling t-SNE
to large data sets.

>>>>

$t$-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method that to aims to find a low-dimensional embedding
where neighborhoods are preserved. More formally, given a multi-dimensional data set
$\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \}
\in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE
aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1,
\mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$,
such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the
high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and
$\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization
tool, $d$ is typically set to two. The similarity between two data points in the
high-dimensional space is defined using the Gaussian kernel
\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
\label{eq:p_ij}
\end{equation}
where $\mathcal{D}$ is some distance measure. The values $p_{j \mid i}$
are then symmetrized to
\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

@@

We introduce \opentsne, a comprehensive Python library that implements the t-SNE algorithm and its numerous recently proposed extensions. \opentsne\ is designed to accommodate all types of users and includes both a simple API that allows users to quickly obtain informative t-SNE visualizations, as well as a more flexible API that allows seasoned practitioners to modify all the
different aspects of the t-SNE algorithm.

The core t-SNE algorithm is comprised of three main steps, mirrored in the
structure of the \opentsne\ library:

\begin{enumerate}
\item Firstly, an affinity model describing the similarities between data points must be specified. Standard t-SNE predominantly deals with tabular data, and similarities are computed using Eqn.~\ref{eq:p_ij}. However, \citet{kobak2019art} showed that using mixtures of Gaussians instead of a single Gaussian often reveals better global consistency of the resulting visualization (albeit at some computational cost). Alternatively, one can speed up the overall computation by choosing a uniform kernel, which yields quantitatively similar results to standard t-SNE. \code{openTSNE.affinity} provides several affinity models to choose from but defaults to the standard perplexity-based model proposed in the original publication.

Furthermore, it is straightforward to define and use a custom affinity model. For instance, one can easily extend the \code{openTSNE.affinity.Affinities} class and provide an affinity model for relational data using a normalized adjacency matrix. This enables rapid experimentation and opens the doors to the visualization of non-tabular data.

\item Secondly, initial positions must be chosen for each data point in the embedding space. These can either be chosen randomly or can be used to impose some initial global structure on the embedding, e.g., using a PCA-based initialization. Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, or the user can also specify their own initialization by providing a \code{numpy.ndarray} object. By default, \opentsne\ opts for PCA-based initialization.

\item Finally, equipped with an affinity model and initial point positions, the only remaining step is to optimize the embedding such that the positions of the points in the embedding space reflect their corresponding similarities in the affinity model. In \opentsne, the initialization and affinities are used to construct an \code{openTSNE.TSNEEmbedding} object, which can then be optimized using different parameter settings. Once an embedding object is constructed, it can subsequently be optimized via the \code{.optimize} method.

Optimization can be performed using either the Barnes-Hut or FIt-SNE approximation scheme. While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, which is often unnecessary for smaller data sets. For smaller data sets, this sometimes results in longer runtimes than when using the Barnes-hut approximation scheme. By default, \opentsne\ uses a heuristic to ensure the fastest algorithm is chosen for the given data set, but this setting can also be specified by the user.
\end{enumerate}

Once the embedding is constructed and optimized, it is ready to use for any downstream tasks. \code{openTSNE.TSNEEmbedding} objects inherit from \code{numpy.ndarray}, which enables easy manipulation and provides interoperability with the broader \proglang{Python} ecosystem.

Embedding new data points to an existing \code{openTSNE.TSNEEmbedding}
follows the same three general steps as constructing a standard embedding:
\begin{enumerate}
\item Firstly, an affinity model must be provided describing the similarities between each new data point and the data points in the reference embedding. As before, \opentsne\ provides several possible affinity models in the \code{openTSNE.affinity} module. Most often, however, we want to utilize the same affinity model that was used to construct the reference embedding, as this leads to more spatially consistent embeddings. This is the default behavior of \opentsne.

\item Secondly, the initial positions of the new data points in the embedding space must be specified. These can, again, be initialized randomly or be informed by their similarity to the data points in the reference embedding. As before, \opentsne\ provides several initialization options in the \code{openTSNE.initialization} module, but the user is free to provide their own initialization.

\item Finally, what remains is to construct an embedding object, which can then be subject to optimization. Because this new embedding relies on an existing reference embedding, we construct a \code{openTSNE.PartialTSNEEmbedding} embedding object, which indicates its dependence on a reference \code{openTSNE.TSNEEmbedding} object. As before, \code{openTSNE.PartialTSNEEmbedding} inherits from \code{numpy.ndarray} and can therefore be easily manipulated. Optimization is, again, performed via the \code{.optimize} method.
\end{enumerate}

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.75\textwidth]{opentsne-diagram}
  \caption{\label{fig:diagram}
  The high-level structure and flow of the \opentsne\ library.
  }
\end{figure*}

Fig.~\ref{fig:diagram} shows the high-level structure and flow of the \opentsne\ library.

Through different parameters, the user is given full control over every aspect of the t-SNE visualization algorithm, from different ways to define affinities to the delicate tuning of optimization parameters. This allows for rapid experimentation and enables users to quickly and efficiently inspect their data from numerous angles and at varying levels of resolution.

While \opentsne\ allows for an unprecedented level of control at every step, most end-users often only need a quick and easy way to create informative visualizations of their data sets. \opentsne\ provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit in the machine learning community. This allows users of all levels to quickly pick up and make use of the \opentsne\ library.

\subsection{Example}
\label{sec:iris}

@@

\opentsne\ is designed to be easy to use and can be installed through either \pkg{PyPI} using \pkg{pip}
\begin{CodeChunk}
\begin{CodeInput}
pip install opentsne
\end{CodeInput}
\end{CodeChunk}
or using \pkg{conda} from the \pkg{conda-forge} channel
\begin{CodeChunk}
\begin{CodeInput}
conda install -c conda-forge opentsne
\end{CodeInput}
\end{CodeChunk}
\opentsne\ provides precompiled binaries for all major platforms (Windows, MacOS, Linux) for all currently supported \proglang{Python} versions (3.7---3.10). This is especially convenient for Windows users, as this obviates the need for a \proglang{C}/\proglang{C++} compiler, which is typically not bundled with the operating system.

@@

We demonstrate the simplest usage of \opentsne\ on the Iris data set from \citet{anderson1936species}. This small data set describes 150 iris flowers containing four variables (sepal length, sepal width, petal length, petal width) and their species (Iris Setosa, Iris Versicolor, Iris Virginica), which is the target variable. This data set is included in \pkg{scikit-learn}:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
\pkg{scikit-learn} loads the data set into a \pkg{numpy} array, which can directly
be used by \opentsne:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
>>>
>>> embedding = openTSNE.TSNE().fit(iris.data)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
Fig.~\ref{fig:iris} shows a visualization of the resulting t-SNE visualization. As expected, t-SNE is correctly able to group together samples from the same species. Additionally, the t-SNE embedding reveals that Iris Setosa is markedly different from the other two species, while flowers from Iris Versicolor and Iris Virginica may appear more similar to one another.
\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{iris}
  \caption{\label{fig:iris}
  A visualization of the obtained t-SNE embedding on the Iris data set. The colors
  correspond to different species of the Iris flower.
}
\end{figure*}
We can now use this embedding as a reference for new data points. In this simple example, we will duplicate some of the original samples and add them into the embedding.
\begin{CodeChunk}
\begin{CodeInput}
>>> new_data = iris.data[::3]
>>> new_embedding = embedding.transform(new_data)
>>> new_embedding[:3]
PartialTSNEEmbedding([[-0.02637324, 15.69956674],
                      [-2.84200792, 16.15478451],
                      [-2.4594136 , 15.37910876]])
\end{CodeInput}
\end{CodeChunk}
Fig.~\ref{fig:iris-transform} shows the resulting positions of the new data points, overlaid the original, reference embedding. As expected, t-SNE is correctly able to place the new data points to their corresponding groups.
\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{iris-transform}
  \caption{\label{fig:iris-transform}
  A visualization of the original, reference data set (transparent points) overlaid with the new data points (opaque points).
}
\end{figure*}

\subsubsection{Peeking under the hood}

While the above example offers users a quick and easy way to generate informative visualizations, \opentsne\ allows us to control the t-SNE algorithm at every step. To demonstrate the flexibility of \opentsne, we will repeat the same analysis of the Iris data set as before, this time leveraging the more advanced features of the library. More practical examples of advanced usage will be shown in Section~\ref{sec:discussion}.

We, again, begin by loading the Iris data set using \pkg{scikit-learn}:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
Next, we import the \opentsne\ library:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
\end{CodeInput}
\end{CodeChunk}
We now mirror the three steps listed above. First, we select an affinity model which describes the similarities between data points:
\begin{CodeChunk}
\begin{CodeInput}
>>> affinities = openTSNE.affinity.PerplexityBasedNN(iris.data, perplexity=30)
\end{CodeInput}
\end{CodeChunk}
Next, we determine the initial point positions in the embedding:
\begin{CodeChunk}
\begin{CodeInput}
>>> initialization = openTSNE.initialization.pca(iris.data)
>>> initialization[:3]
array([[-1.30971087e-04,  1.55848907e-05],
       [-1.32435711e-04, -8.63672049e-06],
       [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Next, using the affinity model and initialization, we can create an embedding object:
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding = openTSNE.TSNEEmbedding(initialization, affinities)
>>> embedding[:3]
TSNEEmbedding([[-1.30971087e-04,  1.55848907e-05],
               [-1.32435711e-04, -8.63672049e-06],
               [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Notice that the embedding object contains the same point coordinates as the initialization. This is because although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization. To optimize the embedding, we merely have to call the \code{embedding.optimize} method using our desired parameter settings. Here, we use the default parameter settings, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding.optimize(250, exaggeration=12, momentum=0.5, inplace=True)
>>> embedding.optimize(500, momentum=0.8, inplace=True)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
We can validate that method produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(openTSNE.TSNE().fit(iris.data) - embedding)
0.0
\end{CodeInput}
\end{CodeChunk}
Next, we can add new samples into the embedding:
\begin{CodeChunk}
\begin{CodeInput}
>>> new_embedding = embedding.prepare_partial(iris.data[::3], perplexity=5)
>>> new_embedding.optimize(
...     exaggeration=1.5, n_iter=250, learning_rate=0.1, momentum=0.8,
...     max_grad_norm=0.25, inplace=True
... )
>>> new_embedding[::3]
PartialTSNEEmbedding([[-0.02637324, 15.69956674],
                      [-2.84200792, 16.15478451],
                      [-2.4594136 , 15.37910876]])
\end{CodeInput}
\end{CodeChunk}
And validate that this also produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(new_embedding - embedding.transform(iris.data[::3]))
0.0
\end{CodeInput}
\end{CodeChunk}

\section{Discussion and Examples} \label{sec:discussion}

@@

We illustrate this point by generating four different embeddings of the data on
single-cell gene expression in mouse brain~\citep{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. We generate
the four different embeddings using the following code snippet (shortened for clarity).
\begin{CodeChunk}
\begin{CodeInput}
embedding_a = openTSNE.TSNE(perplexity=30).fit(x)
embedding_b = openTSNE.TSNE(perplexity=500).fit(x)
embedding_c = openTSNE.TSNE(perplexity=[30, 500]).fit(x)
embedding_d = openTSNE.TSNE(perplexity=30, dof=0.6).fit(x)
\end{CodeInput}
\end{CodeChunk}
While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear equidistant from their neighbors, and the overall relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent. Here, it is evident from the increased white space between groups that there is one large class of excitatory neurons and two related classes of inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods leads to the absorption of smaller clusters into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve both the global organization of clusters and prevent smaller cluster absorption. We constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. The figure demonstrates that some clusters are composed of numerous, smaller subgroups representing different cell subpopulations that are not visible under standard parameter settings.

@@

When dealing with data containing millions of data points, standard t-SNE
embeddings often become unwieldy -- cluster boundaries are blurred, large
clusters absorb smaller ones, and relationships between clusters become
increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from
the data containing expression profiles of over two million single cells
captured at different time points in mouse development. The embedding indicates
numerous clusters with transitions between time points, as marked by the
color-coding, that are difficult to interpret. \citet{kobak2019art} that
increasing attractive forces between similar data points controlled via
the \textit{exaggeration} parameter leads to more compact clusters, and
subsequently, more informative visualizations. For instance, \ref{fig:cao}.b
doubles the default exaggeration, which uncovers some
of the data's overall structure. Further doubling the exaggeration in
\ref{fig:cao}.c allows us to observe that the data is comprised of two main groups
of cells and eight somewhat smaller clusters. The visualization also reveals
several tiny clusters, possibly corresponding to rare cell types.
\begin{CodeChunk}
\begin{CodeInput}
# To improve the global structure of the final layout, we initialize the
# embedding by creating a t-SNE plot on a sample of points

@@

# Create a t-SNE embedding on the sample, using a high perplexity value to
# emphasize the high level structure
init_sample = openTSNE.TSNE(perplexity=500).fit(x_sample)

@@

# Lastly, we merge `init_sample` and `init_rest` and restore the original order
init_full = np.vstack((init_sample, init_rest))[reverse]
\end{CodeInput}
\end{CodeChunk}
We generate the three different embeddings using the following code snippet
(shortened for clarity).
\begin{CodeChunk}
\begin{CodeInput}
# We separately precompute the affinities as this can take quite a long time.
# This way, we can reuse the precomputed affinities in various different
# embeddings, without having to repeat the same, costly computation every time.
affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)

@@

Exaggeration can highlight transitions between cell states in developmental
studies. Standard t-SNE often produces embeddings with clearly defined, discrete
clusters. We can adjust the level of granularity and resolution of the clusters
with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
often undesired in developmental studies where cells' state is assumed to follow
a continuous transition path. To this end, other embedding techniques such as
UMAP and ForceAtlas2 are used to better capture the continuity between cell
states. Recently, \citet{bohm2020unifying} showed that
embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$
construct embeddings that are markedly similar to UMAP and ForceAtlas2,
respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory
between different time points is difficult to observe due to many sprawled out
clusters. On the other hand, it is easier to trace the development when we
increase the exaggeration factor from $1$ to $2$ to $4$ in
Figs.~\ref{fig:cao}.b-c.

@@

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
previously labeled single-cell data set and embed cells from a separate
experiment into the reference landscape. The reference data from
\citet{macosko2015highly} contains gene expression profiles from mouse retinal
cells. By embedding the samples from a similar experiment on bipolar retinal
cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar
cell clusters onto the reference embedding. The following snippet shows the
code necessary to generate figure~\ref{fig:macosko}.
\begin{CodeChunk}
\begin{CodeInput}
# We specify the old default parameter values which are often used to show
# why t-SNE misbehaves or fails to preserve global structure
tsne_old = openTSNE.TSNE(
    initialization="random", learning_rate=200, n_iter=750
).fit(x)

# By default, openTSNE uses PCA-based initialization and automatically
# determines the best learning rate. This also allows us to lower the default
# number of iterations to 500, lowering the overall runtime
tsne_standard = openTSNE.TSNE(perplexity=[50, 500]).fit(x)

@@

Versatility, the ability to use and combine different optimization approaches to
construct different embedding spaces, is another of \opentsne's core design
principles. \citet{kobak2019art} recently provided several recommendations and
tricks to obtain better and more meaningful t-SNE visualizations. These include
multi-scale similarity kernels, perplexity annealing, and increasing
exaggeration when working with massive data sets. \opentsne\ provides a flexible
API to incorporate these improvements in just a few lines of code. Furthermore,
\opentsne\ supports custom affinity models, enabling users to construct t-SNE
embeddings on non-tabular relational data: the only requirement imposed by the
affinity model is some notion of similarity between data points. Finally,
\opentsne's comprehensive callback system can be utilized to monitor and adapt
different stages of the optimization phase and has been used to construct
visually appealing animations of the t-SNE optimization process.

@@

% General problem statement
One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\citep {becht2019dimensionality}. The culprit for slow response time stems from particular optimization procedures and their specific implementation in popular open-source libraries. Until quite recently, most popular implementations of t-SNE were based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation scheme developed by \citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log N)$. Recently, however, \citet{linderman2019fast} developed a novel approximation scheme -- FIt-SNE -- which reduces the asymptotic time complexity to $\mathcal {O}(N)$, overcoming the difficulties of scaling t-SNE to large data sets.

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python~\citep{vanrossum1995python} t-SNE
implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE}
(v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne\ (v0.6.2). Benchmarks were
performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
Benchmarks were run for $1,000$ iterations with the original t-SNE parameters,
as some implementations do not allow for their modification. To ensure a
realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse
brain data set~\citep{xxx}, containing 1.3 million single-cell gene-expression
profiles. We preprocess the data using the standard single-cell analysis
pipeline~\citep{xxx} and extract the top 50 principal components. To generate
benchmark data sets of different sample sizes, we subsample the data set five
times at each sample size. In total, we run each implementation on 30 different
data sets.
